---
title: "vignette1: basic_demo"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{basic_demo}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  echo = FALSE,      # hide code
  message = FALSE,   # hide messages
  warning = FALSE,    # hide warnings
  fig.width = 12,
  fig.height = 7,
  out.width = "100%"
)
```

```{r}
library(gsClusterDetect)
suppressWarnings(suppressMessages(library(data.table)))
data("example_count_data", package = "gsClusterDetect")
data("spline_01", package = "gsClusterDetect")
data("counties", package = "gsClusterDetect")
```

## Overview: Creation of artificial target signals

This package provides a function `st_detect` to create artificial cluster signals for addition to background data in order to test and evaluate the detection capability of the package, particularly the `find_clusters` method. The function adds a total number of case counts to a cylindrical region. The counts are added in the 3-column input frame to distance matrix locations within a fixed radius of central location.

## Available options for signal injections

For purely spatial clusters, the counts are added on a single date, or they may be spread over multiple days. The total number of added case counts, the final injection date, the fixed radius, and the number of inject days are all prescribed by the user as calling arguments to st_inject.

There are two ways to specify the center location. Most directly, the user may use one of the distance matrix locations for the argument `target_loc`. This cluster center need not be included with case counts in the input frame `cases`. The user may also have `st_inject` randomly choose the center location from an area within the distance matrix where counts are dense or sparse. For this option, distance matrix locations are partitioned into deciles according to their total number of counts in the input frame `cases`. Instead of a specific center location, the user may use NULL for center_loc and an integer from 1 to 10 for center_decile, where a value 1 is used to indicate a sparse-data rural location and a value of 10 indicates a dense-count urban location.

## Generating Cluster Signals for Detection

Calling arguments of `st_inject` are:

- cases : a data frame of (date, location, count) triples in the same format
- distance matrix : as in
- target_loc : one of the locations in the distance matrix, or if NULL,
- center_decile : if target_loc is NULL, an integer between 1 and 10
- radius_miles : conceptually, the maximum cluster radius about the center location;
- technically, the largest allowable distance from the center location centroid to the centroid of another location in the cluster
- nr_cases : total number of cases to be injected
- detect_date: final date for the injected cluster
- nr_days: number of days with injected case; if nr_days = 1, all injects are on detect_date

## Generating Cluster Signals for Detection: Example

The following example begins with the package's `example_count_data` table, randomizes counts to ensure no initial clusters, and injects a 100 case counts over a 20-mile radius at a randomly chosen rural location. The `find_clusters` function is applied to illustrate detection.

```{r, echo=TRUE}
# Use example count data to find clusters randomize counts to ensure no initial clusters
d = data.table::copy(example_count_data)
# Randomize counts with Poisson distribution draws to ensure no initial clusters
d[, count:=rpois(n=1,lambda=15), .(location, date)]
detect_date <- d[, max(date)]
# dm <- create_dist_list("county", st = "OH", threshold = 50, )
distance_locations <- county_distance_matrix("OH", unit = "miles")
dm <- distance_locations$distance_matrix
cl <- find_clusters(d, dm, detect_date)  # to confirm absence of clusters

# Now inject a 1-day cluster with a 20-mile radius at a randomly chosen rural location
  set.seed(1)
  nr_cases1 <- 100  
  target_loc <- NULL   # Choose random center location
  center_decile <- 1
  radius_miles <- 20
  nr_days <- 1
  scen1 <- st_injects(d, dm, target_loc, center_decile, 
                      radius_miles, nr_cases1, nr_days, detect_date)
  data_table <- scen1$case_counts_inj
# Get Cluster Data
cluster_data=find_clusters(data_table,dm,detect_date)

# Get tigris based shape file
oh <- tigris::counties("OH", cb = TRUE, class = "sf")
# Pass these to the map_clusters() function
map_clusters(cl = cluster_data, s = oh, s_id = "GEOID")
```

## Example with multiple artificial signals

To create multiple clusters, calls to `st_inject` may be chained as follows:

```{r, echo = TRUE}
nr_cases1 <- 100  # Set = 0 to skip injects  #100
if (nr_cases1 > 0) {
  set.seed(1)
  target_loc <- "39037"  # Darke
  center_decile <- 1
  radius_miles <- 30
  nr_days <- 1
  scen1 <- st_injects(d, dm, target_loc, center_decile, 
                      radius_miles, nr_cases1, nr_days, detect_date)
  d1 <- scen1$case_counts_inj
  nr_cases2 <- 30   # Set = 0 to skip second set of injects  #30
  if(nr_cases2 > 0) {
    target_loc <- "39045"  # Fairfield
    center_decile <- 1
    radius_miles <- 0.1
    nr_days <- 2
    scen2 <- st_injects(d1, dm, target_loc, center_decile, 
                        radius_miles, nr_cases2, nr_days, detect_date)
    data_table <- scen2$case_counts_inj
  } else {
    data_table <- d1
  }
} else {
  data_table <- d
}
# Get Cluster Data
cluster_data=find_clusters(data_table,dm,detect_date)
print(cluster_data$cluster_alert_table)
# Get tigris based shape file
oh <- tigris::counties("OH", cb = TRUE, class = "sf")
# Pass these to the map_clusters() function
map_clusters(cl = cluster_data, s = oh, s_id = "GEOID")
```
