---
title: "Vignette 2: Creating Distance Objects"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette 2: Creating Distance Objects}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  echo = FALSE, # hide code
  message = FALSE, # hide messages
  warning = FALSE, # hide warnings
  fig.width = 12,
  fig.height = 7,
  out.width = "100%"
)
```

```{r setup}
library(gsClusterDetect)
```

## Overview

When using this package to identify clusters, given a set of date-location-count data, one of the key elements that must be provided is the geographical information about the units of clustering. Specifically, we need to provide to the `find_clusters()` function an object that contains information about pairwise distances (See `vignette("basic_demo")` for overview information on finding clusters).

Here we describe how the package provides utilities to generate these objects.

## Types of Distance Objects:

There are two types of distance objects that can be passed to the `find_clusters()` function. These include square pairwise distance matrices and a constrained version that contains only a limited subset of the locations.

### Square pairwise distance matrices

If we have a set of locations among which we would like to identify clusters (say all the counties in Minnesota, or all the zipcodes in Maryland), we can create a matrix containing all the pairwise distances between the centroids of these locations. The package provides functions to do this for census tracts, zipcodes, and counties.

 - `county_distance_matrix()`
 - `zip_distance_matrix()`
 - `tract_distance_matrix()`
 
Each of these functions must be called with a state level parameter, to constrain to a particular state (To combine states, see the section on "Custom Distance Objects"). These functions return a two-element named list containing:

 - `loc_vec`: a vector containing the names of the `N` locations in the geographic target (e.g. "Minnesota")
 - `distance_matrix`: an `N x N` matrix containing the pairwise distances between all locations


### Example: county distance matrix:

```{r eval=TRUE, echo=TRUE}
minnesota_counties <- county_distance_matrix(st = "MN")

# show the length and first 10 values of the `loc_vec`
length(minnesota_counties$loc_vec)
minnesota_counties$loc_vec |> head(10)

# show the dimension and upper left section of the `distance_matrix`
dim(minnesota_counties$distance_matrix)
minnesota_counties$distance_matrix[1:5,1:5]
```

### Example: zipcode distance matrix:

```{r eval=TRUE, echo=TRUE}
maryland_zips <- zip_distance_matrix(st = "MD")

# show the length and first 10 values of the `loc_vec`
length(maryland_zips$loc_vec)
maryland_zips$loc_vec |> head(10)

# show the dimension and upper left section of the `distance_matrix`
dim(maryland_zips$distance_matrix)
maryland_zips$distance_matrix[1:5,1:5]
```

The default unit for these objects is "miles". However, for all of these functions, one can also obtain distances in other units by passing "kilometers" or "meters" to the `unit` parameter.
```{r echo=TRUE, eval=FALSE}
maryland_zip <- zip_distance_matrix(st="MD", unit="kilometers")
```
