---
title: "Vignette 2: Creating Distance Objects"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette 2: Creating Distance Objects}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  echo = FALSE, # hide code
  message = FALSE, # hide messages
  warning = FALSE, # hide warnings
  fig.width = 12,
  fig.height = 7,
  out.width = "100%"
)
```

```{r setup}
library(gsClusterDetect)
```

## Overview

When using this package to identify clusters, given a set of date-location-count data, one of the key elements that must be provided is the geographical information about the units of clustering. Specifically, we need to provide to the `find_clusters()` function an object that contains information about pairwise distances (See `vignette("basic_demo")` for overview information on finding clusters).

Here we describe how the package provides utilities to generate these objects.

## Types of Distance Objects:

There are two types of distance objects that can be passed to the `find_clusters()` function. These include square pairwise distance matrices and a constrained version that contains only a limited subset of the locations.

### Square pairwise distance matrices

If we have a set of locations among which we would like to identify clusters (say all the counties in Minnesota, or all the zipcodes in Maryland), we can create a matrix containing all the pairwise distances between the centroids of these locations. The package provides functions to do this for census tracts, zipcodes, and counties.

 - `county_distance_matrix()`
 - `zip_distance_matrix()`
 - `tract_distance_matrix()`
 
Each of these functions must be called with a state level parameter (`st`), to constrain to a particular state (To combine states, see the section on "Custom Distance Objects"). There is also a `unit` parameter, which takes a string: `"miles"` (default), `"kilometers"`, or "`meters"` to obtains the distance estimates in alternative units. 

All of these functions return a two-element named list containing:

 - `loc_vec`: a vector containing the names of the `N` locations in the geographic target (e.g. "Minnesota")
 - `distance_matrix`: an `N x N` matrix containing the pairwise distances between all locations


#### Example: county distance matrix:

```{r eval=TRUE, echo=TRUE}
minnesota_counties <- county_distance_matrix(st = "MN")

# show the length and first 10 values of the `loc_vec`
length(minnesota_counties$loc_vec)
minnesota_counties$loc_vec |> head(10)

# show the dimension and upper left section of the `distance_matrix`
dim(minnesota_counties$distance_matrix)
minnesota_counties$distance_matrix[1:5,1:5]
```

#### Example: zipcode distance matrix:

```{r eval=TRUE, echo=TRUE}
maryland_zips <- zip_distance_matrix(st = "MD")

# show the length and first 10 values of the `loc_vec`
length(maryland_zips$loc_vec)
maryland_zips$loc_vec |> head(10)

# show the dimension and upper left section of the `distance_matrix`
dim(maryland_zips$distance_matrix)
maryland_zips$distance_matrix[1:5,1:5]
```

#### Example: tract distance matrix:
The package also provides a function to generate distance matrices between census tracts. However, unlike at the zipcode and the county level, there is no built-in dataset that contains the centroids of these tracts across the United States. Therefore, to use the `tract_distance_matrix()` function, the user will need to have the package `tigris` installed. Also, in addition to the `st` and `unit` parameters, the user can also restrict the tracts within a state by providing a vector of 3-character fips codes in the `county` parameter. Furthermore, to prevent the `tigris` package from using a local cache (this has the benefit of speed for repeated calls), the user can turn this cache off using `use_cache = TRUE`

```{r echo=TRUE, eval=TRUE}

cook_county_tracts <- tract_distance_matrix(
  st = "IL",
  county = "031" # the full 5-digit code for Cook County, IL is 17031
)

# show the length and first 10 values of the `loc_vec`
length(cook_county_tracts$loc_vec)
cook_county_tracts$loc_vec |> head(10)

# show the dimension and upper left section of the `distance_matrix`
dim(cook_county_tracts$distance_matrix)
cook_county_tracts$distance_matrix[1:5,1:5]

```

As described above, the default unit for these objects is "miles". However, for all of these functions, one can also obtain distances in other units by passing "kilometers" or "meters" to the `unit` parameter.

```{r echo=TRUE, eval=FALSE}
maryland_zip <- zip_distance_matrix(st="MD", unit="kilometers")
```

### Custom Distance Matrices
Notice that all of the above functions require the user to indicate a state parameter, `st`, thus restricting the output to a single state. We also provide a function `custom_distance_matrix()` to allow users to provide any type of geospatial unit. The function has a different set of parameters than the other functions described above:

 - `df`: This is a `data.frame` with one row per unit, a unique label, and columns containing the latitude and longitude of the centroid of each unit
 - `unit`: as before, this allows the user to obtain distances in miles, kilometers, or meters
 - `label_var`: this is the string name of the column containing the name/label of the unit
 - `lat_var`: the user must provide the name of the column containing the latitude of the centroid
 - `long_var`: the user must provide the name of the column containing the longitude of the centroid

 Below, we demonstrate how this function could be used to get a pairwise distance matrix for a collection of contiguous states, in this case Maryland, Delaware, and Virginia

```{r echo=TRUE, eval=TRUE}
 
 # Use the built-in-counties dataset to get a dataframe of counties in the states of interests
 states <- c("Delaware", "Maryland", "Virginia")
 delmarva_counties <- counties[state_name %in% states]

 head(delmarva_counties, 3)

 # Use the custom function to get the distance matrix
 delmarva_dm <- custom_distance_matrix(
  df = delmarva_counties, 
  label_var = "fips",
  lat_var = "latitude",
  long_var = "longitude"
 )
```

The output structure of the `custom_distance_matrix()` function is the same as the other distance matrix functions.

```{r eval = TRUE, echo=TRUE}

# show the length and first 10 values of the `loc_vec`
length(delmarva_dm$loc_vec)
delmarva_dm$loc_vec |> head(10)

# show the dimension and upper left section of the `distance_matrix`
dim(delmarva_dm$distance_matrix)
delmarva_dm$distance_matrix[1:5,1:5]

```




 
 