---
title: "Vignette 3: Accessing Interim Object from Cluster Estimation" output:
rmarkdown::html_vignette vignette: > %\VignetteIndexEntry{Vignette 3: Accessing
Interim Object from Cluster Estimation} %\VignetteEngine{knitr::rmarkdown}
%\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  echo = FALSE, # hide code
  message = FALSE, # hide messages
  warning = FALSE, # hide warnings
  fig.width = 12,
  fig.height = 7,
  out.width = "100%"
)
```

```{r setup}
library(gsClusterDetect)
```

# Overview

The `find_clusters()` function is the main entry point for this package, and under typical conditions, the function either returns a message indicating that no clusters were found, or returns an object of class "clusters", which is a two element list containing 1) the clusters found and 2) information about the locations in each of those clusters. However, callers can pass `return_interim=True` to obtain the interim or intermediate objects that are estimated under the hood when `find_clusters()` is called.

## Functions called by `find_clusters()`

`find_clusters()` is a wrapper function around a number of other package functions, and it calls the below functions, in order. When the user passes `interim_results=TRUE` the outputs of each of these functions are stored separately within an expanded output list.

+-------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+-------------------------------------------------------------------------------------------------------------------------------------+
| Function                                                                                                                                                                                                                                                                                                    | Output                                                                                                                              |
+=============================================================================================================================================================================================================================================================================================================+=====================================================================================================================================+
| `generate_case_grids()` : this function take the case counts and the detect date (see `?generate_case_grids` for more options) and separates out the counts by baseline and test date interval, and estimates cumulative sums of cases by location.                                                         | An object of class `CaseGrids`, which is a lit of items:                                                                            |
|                                                                                                                                                                                                                                                                                                             |                                                                                                                                     |
|                                                                                                                                                                                                                                                                                                             | - baseline_counts_by_location: a dataframe of counts by location during the baseline interval                                       |
|                                                                                                                                                                                                                                                                                                             |                                                                                                                                     |
|                                                                                                                                                                                                                                                                                                             | - case_grid: a frame of cases during the test period, with reverse cumulative counts within location, by date                       |
|                                                                                                                                                                                                                                                                                                             |                                                                                                                                     |
|                                                                                                                                                                                                                                                                                                             | - case_grid_totals_by_date: a frame of the reverse cumulative sum of counts over all locations, by date                             |
|                                                                                                                                                                                                                                                                                                             |                                                                                                                                     |
|                                                                                                                                                                                                                                                                                                             | - test_cases: a frame of cases during the test period, by and location                                                              |
|                                                                                                                                                                                                                                                                                                             |                                                                                                                                     |
|                                                                                                                                                                                                                                                                                                             | - detect_date: the detect date passed to this function                                                                              |
|                                                                                                                                                                                                                                                                                                             |                                                                                                                                     |
|                                                                                                                                                                                                                                                                                                             | - baseline_total: the total number of cases over all locations and dates                                                            |
+-------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+-------------------------------------------------------------------------------------------------------------------------------------+
| `gen_nearby_case_info()` : This function takes an object of class `CaseGrids` distance matrix, and a distance threshold, and estimates, for each location, the estimated cumulative sum of cases in nearby locations. Two frames for these estimates are provided: one for test period and one for baseline | An object of class `NearbyClusterGrids`, which is a list of two data frames, including:                                             |
|                                                                                                                                                                                                                                                                                                             |                                                                                                                                     |
|                                                                                                                                                                                                                                                                                                             | - "baseline", holding the nearby information for baseline counts                                                                    |
|                                                                                                                                                                                                                                                                                                             |                                                                                                                                     |
|                                                                                                                                                                                                                                                                                                             | - "test", holding the nearby information for the test interval                                                                      |
+-------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+-------------------------------------------------------------------------------------------------------------------------------------+
| `generate_observed_expected()`: This function takes objects of `NearbyClusterGrids` and `CaseGrids` and derives the observed and expected calculation for all target locations                                                                                                                              | An object of class `ObservedExpectedGrid` , which is a frame that contains the observed and expected value for all target locations |
+-------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+-------------------------------------------------------------------------------------------------------------------------------------+

: Table: Functions called by `find_clusters()`

See help via `?<function>` for further details on each of these above functions, including optional and required parameters and examples of how to use them.

## Example Usage:

Note: we set `return_interim` to `TRUE`

```{r}
result <- find_clusters(
  cases = example_count_data,
  distance_matrix = county_distance_matrix("OH")[["distance_matrix"]],
  detect_date = example_count_data[, max(date)],
  distance_limit = 50, 
  return_interim = TRUE
)

# objects included in the output
names(result)
```
