---
title: "Vignette 3: Accessing Interim Object from Cluster Estimation" output:
rmarkdown::html_vignette vignette: > %\VignetteIndexEntry{Vignette 3: Accessing
Interim Object from Cluster Estimation} %\VignetteEngine{knitr::rmarkdown}
%\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  echo = FALSE, # hide code
  message = FALSE, # hide messages
  warning = FALSE, # hide warnings
  fig.width = 12,
  fig.height = 7,
  out.width = "100%"
)
```

```{r setup}
library(gsClusterDetect)
```

# Overview

The `find_clusters()` function is the main entry point for this package, and under typical conditions, the function either returns a message indicating that no clusters were found, or returns an object of class "clusters", which is a two element list containing 1) the clusters found and 2) information about the locations in each of those clusters. However, callers can pass `return_interim=True` to obtain the interim or intermediate objects that are estimated under the hood when `find_clusters()` is called.

## Functions called by `find_clusters()`

`find_clusters()` is a wrapper function around a number of other package functions, and it calls the below functions, in order. When the user passes `interim_results=TRUE` the outputs of each of these functions are stored separately within an expanded output list.

+----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+---------------------------------------------------------------------------------------------------------------+
| Function                                                                                                                                                                                                                                           | Output                                                                                                        |
+====================================================================================================================================================================================================================================================+===============================================================================================================+
| `generate_case_grids()` : this function take the case counts and the detect date (see `?generate_case_grids` for more options) and separates out the counts by baseline and test date interval, and estimate cumulative sums of cases by location  | An object of class `CaseGrids`, which is a lit of items:                                                      |
|                                                                                                                                                                                                                                                    |                                                                                                               |
|                                                                                                                                                                                                                                                    | - baseline_counts_by_location: a dataframe of counts by location during the baseline interval                 |
|                                                                                                                                                                                                                                                    |                                                                                                               |
|                                                                                                                                                                                                                                                    | - case_grid: a frame of cases during the test period, with reverse cumulative counts within location, by date |
|                                                                                                                                                                                                                                                    |                                                                                                               |
|                                                                                                                                                                                                                                                    | - case_grid_totals_by_date: a frame of the reverse cumulative sum of counts over all locations, by date       |
|                                                                                                                                                                                                                                                    |                                                                                                               |
|                                                                                                                                                                                                                                                    | - test_cases: a frame of cases during the test period, by and location                                        |
|                                                                                                                                                                                                                                                    |                                                                                                               |
|                                                                                                                                                                                                                                                    | - detect_date: the detect date passed to this function                                                        |
|                                                                                                                                                                                                                                                    |                                                                                                               |
|                                                                                                                                                                                                                                                    | - baseline_total: the total number of cases over all locations and dates                                      |
+----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+---------------------------------------------------------------------------------------------------------------+
|                                                                                                                                                                                                                                                    |                                                                                                               |
+----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+---------------------------------------------------------------------------------------------------------------+
|                                                                                                                                                                                                                                                    |                                                                                                               |
+----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+---------------------------------------------------------------------------------------------------------------+

: Table: Functions called by `find_clusters()`

## Example Usage:

```{r}
find_clusters(
  cases = example_count_data,
  distance_matrix = county_distance_matrix("OH")[["distance_matrix"]],
  detect_date = example_count_data[, max(date)],
  distance_limit = 50, 
  return_interim = TRUE
)
```
